---
title: 'Create Animated, Combined Plots in R'
date: 'October 2019'
description: "Analyzing sex differences in top lifts at international powerlifting competitions."
output: 
  md_document:
    variant: markdown_github
---
  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, dpi = 300)
```

In this tutorial, I'm going to outline the steps necessary to create an animated, faceted plot in R. Although rare, combining animated plots can be a powerful way to showcase different elements of the same data (as you'll see below).

In this example, I'm using weightlifting data from the International Powerlifting Federation. For the purposes of this tutorial, we'll look at *differences in top lifts by sex*. A faceted, animated plot is a great option because we'd like to observe **the magnitude of these differences** and **how these differences have evolved over time**.

## Environment Setup

These are the packages we'll need to get started. In my case, I use a [custom theme](https://github.com/connorrothschild/cr) I've developed for stylistic purposes. Feel free to instead run `theme_set(theme_minimal())` rather than use my theme!

```{r}
library(ggplot2)
library(tidyverse)
library(ggtext)
library(gifski)
library(gganimate)
library(cr)

set_cr_theme(font = "Proxima Nova")
# theme_set(theme_minimal())
```

## Load and Clean Data

I've already done a lot of the data cleaning for you. If you'd like to follow along, here's the process (or, skip ahead!).

Here' we'll do some minor cleaning and then reshape the three lifts into one column:

```{r}
ipf_lifts <- readr::read_csv("data/ipf_lifts.csv") %>%
  mutate(year = lubridate::year(date))

ipf_lifts_reshape <- ipf_lifts %>% 
  tidyr::pivot_longer(cols = c("best3squat_kg", "best3bench_kg", "best3deadlift_kg"), names_to = "lift") %>% 
  select(name, sex, year, lift, value)
```

For my visualization, I'm only concerned with the *heaviest* lifts from each year:

```{r}
ipf_lifts_maxes <- ipf_lifts_reshape %>% 
  group_by(year, sex, lift) %>% 
  top_n(1, value) %>% 
  ungroup %>% 
  distinct(year, lift, value, .keep_all = TRUE)
```

The first visualization we'll create for the final output is a **dumbbell plot**. Curious what that is, or how to make it in R? [Check out my other post on the topic](https://www.connorrothschild.com/post/dumbbell-plots/).

In order to construct a dumbbell plot, we need both male and female
observations in the same row. For this, we use the `spread` function.

```{r}
max_pivot <- ipf_lifts_maxes %>% 
  spread(sex, value)
```

Let's try to construct a dataframe for each sex:
  
```{r}
male_lifts <- max_pivot %>% 
  select(-name) %>% 
  filter(!is.na(M)) %>% 
  group_by(year, lift) %>% 
  summarise(male = mean(M))

female_lifts <- max_pivot %>% 
  select(-name) %>% 
  filter(!is.na(`F`)) %>% 
  group_by(year, lift) %>% 
  summarise(female = mean(`F`))
```

And join them:
  
```{r}
max_lifts <- merge(male_lifts, female_lifts)

max_lifts_final <- max_lifts %>% 
  group_by(year, lift) %>% 
  mutate(diff = male - female)
```

Not following along, or want to check your progress? Here's what our data looks like in its final form:

```{r}
max_lifts_final %>% 
  reactable::reactable()
```

## Visualize

Finally, we can construct the visualization.

First, we can create a static visualization using {ggalt} (again, my [blog post](https://www.connorrothschild.com/post/dumbbell-plots/) covers the details of this step).

```{r}
max_lifts_final %>% 
  filter(year == 2019) %>% 
  ggplot() + 
  ggalt::geom_dumbbell(aes(y = lift,
                    x = female, xend = male),
                colour = "grey", size = 5,
                colour_x = "#D6604C", colour_xend = "#395B74") +
  labs(y = element_blank(),
       x = "Top Lift Recorded (kg)",
       title =  "How <span style='color:#D6604C'>Women</span> and <span style='color:#395B74'>Men</span> Differ in Top Lifts",
                       subtitle = "In 2019") +
  theme(plot.title = element_markdown(lineheight = 1.1, size = 20),
        plot.subtitle = element_text(size = 15)) +
  scale_y_discrete(labels = c("Bench", "Deadlift", "Squat")) +
  drop_axis(axis = "y") +
  geom_text(aes(x = female, y = lift, label = paste(female, "kg")),
            color = "#D6604C", size = 4, vjust = -2) +
  geom_text(aes(x = male, y = lift, label = paste(male, "kg")),
            color = "#395B74", size = 4, vjust = -2) +
  geom_rect(aes(xmin=430, xmax=470, ymin=-Inf, ymax=Inf), fill="grey80") +
  geom_text(aes(label=diff, y=lift, x=450), size=4) +
  geom_text(aes(x=450, y=3, label="Difference"),
            color="grey20", size=4, vjust=-3)
```

Finally, we animate, using Thomas Pedersen’s wonderful [gganimate
package](https://github.com/thomasp85/gganimate). This is a relatively easy step, because `gganimate` only requires two extra lines of code: `transition_states` and `ease_aes`. Then, we pass it into the `animate` function!

```{r}
animation <- max_lifts_final %>% 
  ggplot() + 
  ggalt::geom_dumbbell(aes(y = lift,
                    x = female, xend = male),
                colour = "grey", size = 5,
                colour_x = "#D6604C", colour_xend = "#395B74") +
  labs(y = element_blank(),
       x = "Top Lift Recorded (kg)",
       title =  "How <span style='color:#D6604C'>Women</span> and <span style='color:#395B74'>Men</span> Differ in Top Lifts",
  subtitle='\nThis plot depicts the difference between the heaviest lifts for each sex at International Powerlifting Federation\nevents over time. \n \n{closest_state}') +
  theme(plot.title = element_markdown(lineheight = 1.1, size = 25, margin=margin(0,0,0,0)),
        plot.subtitle = element_text(size = 15, margin=margin(8,0,-30,0))) +
  scale_y_discrete(labels = c("Bench", "Deadlift", "Squat")) +
  drop_axis(axis = "y") +
  geom_text(aes(x = female, y = lift, label = paste(female, "kg")),
            color = "#D6604C", size = 4, vjust = -2) +
  geom_text(aes(x = male, y = lift, label = paste(male, "kg")),
            color = "#395B74", size = 4, vjust = -2) +
  transition_states(year, transition_length = 4, state_length = 1) +
  ease_aes('cubic-in-out')

a_gif <- animate(animation, 
                 fps = 10, 
                 duration = 25,
                 width = 800, height = 400, 
                 renderer = gifski_renderer("outputs/animation.gif"))

a_gif
```

But in our case, we'd like to include another GIF: a line chart of differences over time.

```{r}
animation2 <- max_lifts_final %>% 
  ungroup %>% 
  mutate(lift = case_when(lift == "best3bench_kg" ~ "Bench",
                          lift == "best3squat_kg" ~ "Squat",
                          lift == "best3deadlift_kg" ~ "Deadlift")) %>% 
  ggplot(aes(year, diff, group = lift, color = lift)) + 
  geom_line(show.legend = FALSE) + 
  geom_segment(aes(xend = 2019.1, yend = diff), linetype = 2, colour = 'grey', show.legend = FALSE) + 
  geom_point(size = 2, show.legend = FALSE) + 
  geom_text(aes(x = 2019.1, label = lift, color = "#000000"), hjust = 0, show.legend = FALSE) + 
  drop_axis(axis = "y") +
  transition_reveal(year) +
  coord_cartesian(clip = 'off') +
  theme(plot.title = element_text(size = 20)) +
  labs(title = 'Difference over time',
       y = 'Difference (kg)',
       x = element_blank()) + 
  theme(plot.margin = margin(5.5, 40, 5.5, 5.5))

b_gif <- animate(animation2, 
                 fps = 10, 
                 duration = 25,
        width = 800, height = 200, 
        renderer = gifski_renderer("outputs/animation2.gif"))

b_gif
```

Finally, we'll combine them using `magick` (thanks to [this
post](https://github.com/thomasp85/gganimate/wiki/Animation-Composition)):

```{r}
library(magick)
a_mgif <- image_read(a_gif)
b_mgif <- image_read(b_gif)

new_gif <- image_append(c(a_mgif[1], b_mgif[1]), stack = TRUE)
for(i in 2:250){
  combined <- image_append(c(a_mgif[i], b_mgif[i]), stack = TRUE)
  new_gif <- c(new_gif, combined)
}
```

What’s happening here? Essentially, we’re using the power of `magick` to:

1. Read in all of the *individual images* (`image_read`) from each GIF (after all, a GIF is just a series of images!).
2. For the first frame, stack the two images on top of each other (`image_append`), so plot 1 is above plot 2.
3. For the rest of the frames (in my case, the next 249, because my GIF had 250 frames), replicate this and combine it with the first frame (this is the `for` loop).

Here, we specify `stack = TRUE` so that one plot is **above** the other. If you'd like to place them side-by-side, specify `stack = FALSE`.

In combination, the process results in our final output:

```{r}
new_gif
```

In this view, we can see the magnitude of the differences both relatively and absolutely (top chart), *and* we can see how these differences change over time (bottom chart). The power of an animated, combined chart!